String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-30 2023-1-19/sections/overview.tex
overview

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Finally, The satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut\equiv \aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, abbreviated as $\sf NE_{\varphi}(\aut)$. That is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut$, producing resulting registers values $(v_1, v_2, v_3, v_4)$ that satisfy $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $\varphi$. 
It is not hard to observe that $\sf NE_{\varphi}(\aut)$ is independent of the characters of $\aut$.  Therefore, the characters in $\aut$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$, where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut$ (see Figure~\ref{fig:overview:product:reduced}(a)). Then $\sf NE_{\varphi}(\aut)$ is reduced to the problem of deciding whether there exists $w' \in \costset^*$ such that it is accepted by $\cB$ and its Parikh image (i.e., numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies that $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60$ and $v'_4 > 120$, where $(v'_1, v'_2, v'_3, v'_4) =  \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$ for each $\myvec{v}\in \costset$. Intuitively, $(v'_1, v'_2, v'_3, v'_4)$ is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

@@

Nevertheless, the algorithm illustrated above is insufficient when regexes are complicated. The size of $\cB$ can still be big, causing the resulted existential LIA formulas are of big sizes and hard to be solved (NP-complete \cite{Haase18}). 
To tackle this issue, we propose techniques to reduce the sizes of $\cB$ over the alphabet $\costset$.

@@

Specifically, to reduce the sizes of the $\cB$, we determinize $\cB$ and apply the minimization algorithm to the resulting deterministic finite automaton (DFA), obtaining a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b).
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
The evaluation in section \cite{sec} shows that we can solve 1503 more instances after utilizing simplification techniques, with \%54.8 less average time.

We implement all the aforementioned techniques on the top of OSTRICH, resulting in a solver $\ostrichrecl$. It turns out that $\ostrichrecl$ is able to solve the string constraint in~(\ref{eqn-running}) within one second, while the state-of-the-art string solvers are incapable of solving it within 300 seconds.

>>>>

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Finally, the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut\equiv \aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut$, so that the resulting registers values $(v_1, v_2, v_3, v_4)$ satisfy both $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $\varphi$. 
It is not hard to observe that the nonemptiness of $\aut$ with respect to $\varphi$ is independent of the characters of $\aut$.  Therefore, the characters in $\aut$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$, where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut$ (see Figure~\ref{fig:overview:product:reduced}(a)). Then the original problem is reduced to the problem of deciding whether there exists a string $w' \in \costset^*$ that is accepted by $\cB$ and its Parikh image (i.e., numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60 \wedge v'_4 > 120$, where $(v'_1, v'_2, v'_3, v'_4) =  \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$ for each $\myvec{v}\in \costset$. Intuitively, $(v'_1, v'_2, v'_3, v'_4)$ is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

@@

Nevertheless, when the original regexes are complicated (e.g. contains the occurrences of negation or intersection operators), the sizes of the NFA $\cB$ can still be big and the sizes of the LIA formulas defining the Parikh image of $\cB$ are also big. Since the satisfiability of LIA formulas is an NP-complete problem \cite{Haase18}, big sizes of LIA formulas would be a bottleneck of the efficiency. 
To tackle this issue, we propose techniques to reduce the sizes of the NFA $\cB$.
%over the alphabet $\costset$.

@@

Specifically, to reduce the sizes of $\cB$, we take $\vec{0}$-labeled transitions as $\epsilon$-transitions, determinize $\cB$, and apply the minimization algorithm to the resulting deterministic finite automaton (DFA), resulting in a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b).
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
%The evaluation in Table~\ref{tab:results_simp} shows that we can solve 1503 more instances after utilizing simplification techniques, with \%54.8 less average time.

We implement all the aforementioned techniques on the top of OSTRICH, resulting in a solver $\ostrichrecl$. It turns out that $\ostrichrecl$ is able to solve the string constraint in~(\ref{eqn-running}) within one second, while the state-of-the-art string solvers are incapable of solving it within 120 seconds.
